Social Participation Ontology: community documentation, enhancements and use examples

### 2015-09-18 plosOne/plos.tex
adiantada adequacao ao template

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.4\textwidth]{figs/logoUnificadoDoPDF}
    \caption{Some of the various logos for of the \vcps. (a) is a colored text logo proposal; (b) is a figurative logo; (c) is mixture of both ideas. It can be seen that these logos were conceived for the ontology when it was called a \emph{vocabulary}. Community documents reflect this nomenclature, which is being changed with this article and subsequent work with \ops\ advent.}
    \label{logo}
\end{figure}

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.9\textwidth]{figs/opsBFO___}
    \caption{Diagram representation of the Ontology of Social Participation (\ops). Arrows with white heads indicate ``is a'' relations (subclass points to superclass). Arrows with black heads indicate property relations from subject to object . All property relation yield existential restrictions, with the exception of the ``has characteristic'' property, that does not yield restriction. Upper ontologies \bfo\ and \foaf\ are asserted through color (\bfo) and dashed boxes (\foaf).}
    \label{fig:v1}
\end{figure*}

@@

\begin{table*}[!h]
    \footnotesize
  \centering
  \caption{Classes of the \ops\ (Ontology of Social Participation). These are core concepts in the ontology. Along with the taxonomic structure exposed in Figure~\ref{fig:owlCC}, these classes are related by the properties in Table~\ref{ospProps}. Noteworthy is \vcps\ class name ``{\tt vcps:Espa\c{c}o de A\c{c}\~ao}'', which caused original \vcps\ \owl\ to be pointed as corrupted by \protege\ (spaces are not allowed in \uri s). Also, class {\tt vcps:Role} was dropped as it yields logical problens in \vcps.}
  \begin{tabular}{|p{1.8cm}|p{1.6cm}||p{2.2cm}|p{2.2cm}|p{1.8cm}||p{4cm}||p{3cm}|}\hline
      {\bf OPS class name} & {\bf VCPS class name} & {\bf pt-br label} & {\bf es label} & {\bf en label} & {\bf definition} & {\bf upper ontology classes} \\\hline\hline
      Person & Pessoa & Pessoa & Persona & Person & a person (social actor is a person) & {\tt bfo:'Material Entity'}, {\tt foaf:Person} \\ \hline
      Organization & Organizacao & Organiza\c{c}\~ao & Organizaci\'on & Organization & social actor is a group of individuals, organized formally or informally (e.g. movements, collectives) & {\tt bfo:'Material Entity'}, {\tt foaf:Organization} \\ \hline \hline
Executor & Executor & Executor & Ejecutor & Executor & performs action directly and is responsible for results & {\tt bfo:'Material Entity'} \\ \hline
Initiator & Iniciador & Iniciador & Iniciador & Initiator & originates cause, individually or collaborativelly & {\tt bfo:'Material Entity'} \\ \hline
      Supporter & Apoiador & Apoiador & Apoyador & Supporter & supports cause with resources of any kind (e.g. cognitive, financial, equipments) & {\tt bfo:'Material Entity'} \\ \hline
SocialActor & Ator & Ator Social & Actor Social & Social Actor & entity that might have a participatory role & {\tt bfo:'Material Entity'} \\ \hline \hline
Participation-Characteristic & NivelDe-Participacao & Caracter\'istica de Participa\c{c}\~ao & Caracter\'istica de Participaci\'on & Participation Characteristic & the way the participation of the specific actor is happening & {\tt bfo:'Dependent Continuant'} \\\hline
Cause & Causa & Causa & Causa &  Cause & the motivation for Action & {\tt bfo:'Dependent Continuant'} \\\hline
      Scope &{\bf Espa\c{c}o de A\c{c}\~ao} & Escopo & Ambito & Scope & the scope os Action & {\tt bfo:'Dependent Continuant'} \\\hline
Result & Resultados & Resultado & Resultado & Result & the result obtained with action & {\tt bfo:'Dependent Continuant'} \\\hline
      Solution & Solucao & Solu\c{c}\~ao & Soluci\'on & Solution & solution achieved with Action & {\tt bfo:'Dependent Continuant'} \\\hline \hline
      Problem & Problema & Problema & Problema & Problem & the problem that the Action aims to solve & {\tt bfo:'Independent Continuant'} \\\hline
Theme & Tema & Tema & Tema & Theme & the theme in focus by Action & {\tt bfo:'Independent Continuant'} \\\hline \hline
Action & Acao & A\c{c}\~ao & Acci\'on & Action & what is done in terms os social participation & {\tt bfo:'Processual Entity'} \\\hline\hline
      {\color{red} dropped class}& Papel & -//-  & -//- & -//- & the role of the actor & -//- \\ \hline
  \end{tabular}
  \label{ospClasses}
\end{table*}

\begin{table*}[!h]
  \centering
  \caption{Properties of the \ops\ (Ontology of Social Participation) along original \vcps\ names. Only a few ranges were established, and no domain, as these were not currently useful and, without them, \ops\ can be used more freely.}
  \begin{tabular}{|l|l||p{2.2cm}|p{2.2cm}|p{1.8cm}||l|p{2.2cm}|}\hline
{\bf OSP property name} & {\bf CVSP property name} & {\bf pt-br label} & {\bf es label} & {\bf en label} & {\bf domain} & {\bf range} \\\hline\hline
theme & possuiTemaAssociado & tema & tema & theme & -//- & Theme \\ \hline
belongsTo & pertenceAoEspaco & pertence ao & pertence al & belongs to & -//- & Scope \\ \hline
action & possuiAcao & a\c{c}\~ao & acci\'on & action & -//- & Action \\\hline\hline
supports & apoiaCausa & apoia & apoya & supports & -//- & -//- \\ \hline
contributesTo & compoeSolucao & contribui para & contribuye para & contributes to & -//- & -//- \\ \hline
executes & executaAcao & executa & ejecuta & executes & -//- & -//- \\ \hline
generates & geraCausa & gera & genera & generates & -//- & -//- \\ \hline
starts & iniciaCausa & inicia & inicializa & starts & -//- & -//- \\ \hline
solves & soluciona & soluciona & resuelve & solves & -//- & -//- \\ \hline
produces & produzResultado & produz & produce & produces & -//- & -//- \\\hline
proposes & propoeSolucao & prop\~oe & propone & proposes & -//- & -//- \\\hline
      trait & temNivelDeParticipacao & tra\c{c}o & rasgo & trait & -//- & -//- \\\hline\hline
{\color{red} dropped} & temPapel & -//-  & -//-  & -//-  & -//-  & -//- \\ \hline
  \end{tabular}
  \label{ospProps}
\end{table*}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figs/opsTax}
    \caption{A taxonomic tree of the Ontology of Social Participation (\ops). This image was rendered inside \protege, with the \owl\ code in Appendix~\ref{owl:vcps}. Figure~\ref{fig:v1} is more informative, but this diagram is more standard and might be simpler for the newcomer. Note that the taxonomic tree does not present any information about properties further linking these classes. The prefixes {\tt snap:} and {\tt span:} are \bfo\ specific for {\tt snap}shot and {\tt span}ning-time.}
    \label{fig:owlCC}
\end{figure*}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=\columnwidth]{figs/pubbyDer}
    \caption{Dereferencing an \ops\ class: the \uri\ is also an \urll, which, reached by \http, returns information for the user as shown. Also, if the client is not a browser, but a crawler or a linked data application, \pubby\ delivers plain \rdf, not user-friendly \html.}
    \label{fig:deref}
\end{figure}

@@

\begin{figure}
    \centering
        \includegraphics[width=0.9\columnwidth]{figs/endpoint}
    \caption{Scheme of the common use of ontologies for multiple databases integration. A user or application reaches a \sparql\ endpoint. This endpoint, through ontologies, delivers data from one or more databases. Nowadays, the most usual is to find only one database available at an endpoint, and this database is usually duplicated and not synchronized, but available as a (converted) triplestore. Even so, it is possible to access multiple ontologies and it is desirable that the databases have synchronous access, i.e. without need to convert data to triples beforehand.}
    \label{endpoint}
\end{figure}

@@

\begin{sidewaysfigure*}
    \vspace{24.5em}
    \hspace*{-3.5em}
        \includegraphics[width=1.05\textwidth]{figs/opsExpandedTax}
        \caption{A taxonomic diagram of an example expanded instance of \ops. {\tt ops:PaidExecutor} is defined as being the subject of a \texttt{ops:receivesFrom} relation with a \texttt{ops:SocialActor}; \texttt{ops:DownloadedMob} is defined by being the subject of a \texttt{ops:convoquedBy} relation with a \texttt{ops:SocialNetwork}. \owl\ code is online for live editing~\cite{owlExp}. The new classes added to \ops\ are in Table~\ref{ospFooClass}}
    \label{fig:owlExp}
\end{sidewaysfigure*}

>>>>

\begin{figure}[!h]
    \centering
%    \includegraphics[width=0.4\textwidth]{figs/logoUnificadoDoPDF}
    \caption{Some of the various logos for of the \vcps. (a) is a colored text logo proposal; (b) is a figurative logo; (c) is mixture of both ideas. It can be seen that these logos were conceived for the ontology when it was called a \emph{vocabulary}. Community documents reflect this nomenclature, which is being changed with this article and subsequent work with \ops\ advent.}
    \label{logo}
\end{figure}

@@

\begin{figure*}
    \centering
%    \includegraphics[width=0.9\textwidth]{figs/opsBFO___}
    \caption{Diagram representation of the Ontology of Social Participation (\ops). Arrows with white heads indicate ``is a'' relations (subclass points to superclass). Arrows with black heads indicate property relations from subject to object . All property relation yield existential restrictions, with the exception of the ``has characteristic'' property, that does not yield restriction. Upper ontologies \bfo\ and \foaf\ are asserted through color (\bfo) and dashed boxes (\foaf).}
    \label{fig:v1}
\end{figure*}

@@

\begin{table*}[!h]
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
    \footnotesize
  \centering
  \caption{Classes of the \ops\ (Ontology of Social Participation). These are core concepts in the ontology. Along with the taxonomic structure exposed in Figure~\ref{fig:owlCC}, these classes are related by the properties in Table~\ref{ospProps}. Noteworthy is \vcps\ class name ``{\tt vcps:Espa\c{c}o de A\c{c}\~ao}'', which caused original \vcps\ \owl\ to be pointed as corrupted by \protege\ (spaces are not allowed in \uri s). Also, class {\tt vcps:Role} was dropped as it yields logical problens in \vcps.}
  \begin{tabular}{|p{1.8cm}|p{1.6cm}||p{2.2cm}|p{2.2cm}|p{1.8cm}||p{4cm}||p{3cm}|}\hline
      {\bf OPS class name} & {\bf VCPS class name} & {\bf pt-br label} & {\bf es label} & {\bf en label} & {\bf definition} & {\bf upper ontology classes} \\\hline\hline
      Person & Pessoa & Pessoa & Persona & Person & a person (social actor is a person) & {\tt bfo:'Material Entity'}, {\tt foaf:Person} \\ \hline
      Organization & Organizacao & Organiza\c{c}\~ao & Organizaci\'on & Organization & social actor is a group of individuals, organized formally or informally (e.g. movements, collectives) & {\tt bfo:'Material Entity'}, {\tt foaf:Organization} \\ \hline \hline
Executor & Executor & Executor & Ejecutor & Executor & performs action directly and is responsible for results & {\tt bfo:'Material Entity'} \\ \hline
Initiator & Iniciador & Iniciador & Iniciador & Initiator & originates cause, individually or collaborativelly & {\tt bfo:'Material Entity'} \\ \hline
      Supporter & Apoiador & Apoiador & Apoyador & Supporter & supports cause with resources of any kind (e.g. cognitive, financial, equipments) & {\tt bfo:'Material Entity'} \\ \hline
SocialActor & Ator & Ator Social & Actor Social & Social Actor & entity that might have a participatory role & {\tt bfo:'Material Entity'} \\ \hline \hline
Participation-Characteristic & NivelDe-Participacao & Caracter\'istica de Participa\c{c}\~ao & Caracter\'istica de Participaci\'on & Participation Characteristic & the way the participation of the specific actor is happening & {\tt bfo:'Dependent Continuant'} \\\hline
Cause & Causa & Causa & Causa &  Cause & the motivation for Action & {\tt bfo:'Dependent Continuant'} \\\hline
      Scope &{\bf Espa\c{c}o de A\c{c}\~ao} & Escopo & Ambito & Scope & the scope os Action & {\tt bfo:'Dependent Continuant'} \\\hline
Result & Resultados & Resultado & Resultado & Result & the result obtained with action & {\tt bfo:'Dependent Continuant'} \\\hline
      Solution & Solucao & Solu\c{c}\~ao & Soluci\'on & Solution & solution achieved with Action & {\tt bfo:'Dependent Continuant'} \\\hline \hline
      Problem & Problema & Problema & Problema & Problem & the problem that the Action aims to solve & {\tt bfo:'Independent Continuant'} \\\hline
Theme & Tema & Tema & Tema & Theme & the theme in focus by Action & {\tt bfo:'Independent Continuant'} \\\hline \hline
Action & Acao & A\c{c}\~ao & Acci\'on & Action & what is done in terms os social participation & {\tt bfo:'Processual Entity'} \\\hline\hline
      {\color{red} dropped class}& Papel & -//-  & -//- & -//- & the role of the actor & -//- \\ \hline
  \end{tabular}
  \label{ospClasses}
\end{adjustwidth}
\end{table*}

\begin{table*}[!h]
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
%  \centering
  \caption{Properties of the \ops\ (Ontology of Social Participation) along original \vcps\ names. Only a few ranges were established, and no domain, as these were not currently useful and, without them, \ops\ can be used more freely.}
  \begin{tabular}{|l|l||p{2.2cm}|p{2.2cm}|p{1.8cm}||l|p{1.2cm}|}\hline
{\bf OSP property name} & {\bf CVSP property name} & {\bf pt-br label} & {\bf es label} & {\bf en label} & {\bf domain} & {\bf range} \\\hline\hline
theme & possuiTemaAssociado & tema & tema & theme & -//- & Theme \\ \hline
belongsTo & pertenceAoEspaco & pertence ao & pertence al & belongs to & -//- & Scope \\ \hline
action & possuiAcao & a\c{c}\~ao & acci\'on & action & -//- & Action \\\hline\hline
supports & apoiaCausa & apoia & apoya & supports & -//- & -//- \\ \hline
contributesTo & compoeSolucao & contribui para & contribuye para & contributes to & -//- & -//- \\ \hline
executes & executaAcao & executa & ejecuta & executes & -//- & -//- \\ \hline
generates & geraCausa & gera & genera & generates & -//- & -//- \\ \hline
starts & iniciaCausa & inicia & inicializa & starts & -//- & -//- \\ \hline
solves & soluciona & soluciona & resuelve & solves & -//- & -//- \\ \hline
produces & produzResultado & produz & produce & produces & -//- & -//- \\\hline
proposes & propoeSolucao & prop\~oe & propone & proposes & -//- & -//- \\\hline
      trait & temNivelDeParticipacao & tra\c{c}o & rasgo & trait & -//- & -//- \\\hline\hline
{\color{red} dropped} & temPapel & -//-  & -//-  & -//-  & -//-  & -//- \\ \hline
  \end{tabular}
  \label{ospProps}
\end{adjustwidth}
\end{table*}

@@

\begin{figure*}
    \centering
%    \includegraphics[width=\textwidth]{figs/opsTax}
    \caption{A taxonomic tree of the Ontology of Social Participation (\ops). This image was rendered inside \protege, with the \owl\ code in Appendix~\ref{owl:vcps}. Figure~\ref{fig:v1} is more informative, but this diagram is more standard and might be simpler for the newcomer. Note that the taxonomic tree does not present any information about properties further linking these classes. The prefixes {\tt snap:} and {\tt span:} are \bfo\ specific for {\tt snap}shot and {\tt span}ning-time.}
    \label{fig:owlCC}
\end{figure*}

@@

\begin{figure}[!h]
    \centering
%    \includegraphics[width=\columnwidth]{figs/pubbyDer}
    \caption{Dereferencing an \ops\ class: the \uri\ is also an \urll, which, reached by \http, returns information for the user as shown. Also, if the client is not a browser, but a crawler or a linked data application, \pubby\ delivers plain \rdf, not user-friendly \html.}
    \label{fig:deref}
\end{figure}

@@

\begin{figure}
    \centering
%        \includegraphics[width=0.9\columnwidth]{figs/endpoint}
    \caption{Scheme of the common use of ontologies for multiple databases integration. A user or application reaches a \sparql\ endpoint. This endpoint, through ontologies, delivers data from one or more databases. Nowadays, the most usual is to find only one database available at an endpoint, and this database is usually duplicated and not synchronized, but available as a (converted) triplestore. Even so, it is possible to access multiple ontologies and it is desirable that the databases have synchronous access, i.e. without need to convert data to triples beforehand.}
    \label{endpoint}
\end{figure}

@@

%\begin{sidewaysfigure*}
    \begin{figure}
%    \vspace{24.5em}
%    \hspace*{-3.5em}
%        \includegraphics[width=1.05\textwidth]{figs/opsExpandedTax}
        \caption{A taxonomic diagram of an example expanded instance of \ops. {\tt ops:PaidExecutor} is defined as being the subject of a \texttt{ops:receivesFrom} relation with a \texttt{ops:SocialActor}; \texttt{ops:DownloadedMob} is defined by being the subject of a \texttt{ops:convoquedBy} relation with a \texttt{ops:SocialNetwork}. \owl\ code is online for live editing~\cite{owlExp}. The new classes added to \ops\ are in Table~\ref{ospFooClass}}
    \label{fig:owlExp}
    \end{figure}
%\end{sidewaysfigure*}
